---
title: "Data wrangling"
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
    css: ../resources/style.css
  pdf_document:
    df_print: default
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  word_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    toc: yes
    toc_depth: 2
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
---

```{r setup, include=FALSE, warnings=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparations

Load the necessary libraries

```{r libraries, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}
library(tidyverse) #for data wrangling
```

Link to the data transformation cheatsheet

https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf

Important data manipulation functions:

| Task                      | Function                    | Package   |
|---------------------------|-----------------------------|-----------|
| Sorting                   | `arrange()`                 | **dplyr** |
| Re-ordering factor levels | `factor()`                  | **base**  |
| Re-labelling              | `recode()`                  | **dplyr** |
| Re-naming columns         | `rename()`                  | **dplyr** |
| Filtering/subsetting      | `select()`                  | **dplyr** |
| Transformations           | `mutate()`                  | **dplyr** |
| Adding columns            | `mutate()`                  | **dplyr** |
| Re-shaping data           | `gather()`/`spread()`       | **tidyr** |
| Aggregating               | `summarize()`, `group_by()` | **dplyr** |
| Merging/joining           | `*_join()`                  | **dplyr** |
|                           |                             |           |

# Piping

strings single functions together so it performs all these outputs together in sequence
e.g.,



# Data files

```{r getData, results='markdown', eval=TRUE}
load(file='../data/manipulationDatasets.RData')
data.1 %>% head
summary(data.1)
str(data.1)
glimpse(data.1)
```

Sorting data
================

```{r sorting}
arrange(data.1, LAT)

data.1 %>% arrange(LAT)

arrange(data.1, -LAT) # descending order

arrange(data.1, Cond,Temp) # Sort by different things, categorial values presented in alphabetical order
data.1 %>% arrange(Cond & Temp) #same but piped

arrange(data.1, Between, Cond)
```

Manipulating factors
=======================
```{r}
levels(data.1$Cond) # Show what order these factors are in

data.3 <- data.1
levels(data.3$Cond)

data.3$Cond <- factor(data.3$Cond, levels=c('L', 'M', 'H'), labels=c("Low", "Medium", "High")) # Make factors more sensible, order them, give labels

```


Subset columns
=================

```{r}
data.1 %>% select(Between, Plot, Cond, Time, Temp) ##Subset columns

dplyr::select(data.1 , Between, Plot, Cond, Time, Temp)

select(data.1, -LAT, -LONG)

head(data.1)

##helper functions

contains()
ends_with()
starts_with()
matches()

select(data.1, contains('L'))

select(data.1, matches('^T[a-z]m.')) ## Search for start with capital T, then contain a-z, then contain "m", then contain
                                     ## any character


##Practise with nasa data set - select only lat long and anything that starts with cloud
head(nasa)

head(select(nasa, lat, long, starts_with("cloud")))

##everything function can be helpful when sorting
select(data.1, Time, everything())

## use tikus data set
tikus[1:10, c(1:3, 76:77)]

## select rep,time and only species that do not contain pora

dplyr::select(tikus, -contains('pora'))

or

dplyr::select(tikus, rep, time, everything(), -contains('pora'))

##If a variable contains space you have to refer to it with backticks ''
dplyr::select(tikus, 'Psammocora contigua')

##pull spits out individual vector
pull(data.1, Temp)

## rename your columns
rename(data.1, Condition=Cond, Temperature=Temp)
```

## Regular expressions (regex)

https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf


Filtering
=============

```{r}
filter(data.1, Cond=='H') ## Two equal signs means questions, not is equal too, but DOES cond equal H??

filter(data.1, Cond %in% c('H', 'M')) ## for multiple filter criteria

filter(data.1, Cond=='H' & Temp<25) ## cond is h and temp is less than 25

filter(data.1, Cond=='H' & Temp<25) ## cond is h OR temp is less than 25

filter(data.1, Temp<20 & (LAT>20 | LONG<145))

## above is filtering by criteria
## We can also filter by location

slice(data.1, 1:4)

slice(data.1, c(1:4,7))


data.3<-filter(data.1, Plot=='P1') ## thise filters but in graphs and tables still includes empty spaces for dropped variables

##use drop levels

data.1 %>% filter(Plot=="P1") %>% droplevels()

## Always remember to store your subsets to actually work with them

data.processed <- data.1 %>% filter(Plot=="P1") %>% droplevels()

```

Adding columns - mutate
===========================

```{r}
##nonsense but for examples sake

mutate(data.1, LL=LAT+LONG)

##log transform

mutate(data.1, logTemp=log(Temp))

##centering your data
data.1 %>% mutate(cTemp=Temp-mean(Temp))


```

Summarising (aggregating) data
=================================

```{r}
data.1 %>% summarize(Mean=mean(Temp), Var=var(Temp)) ## summary stats

summarise(data.1, across(c(Temp, LAT), list(Mean=mean, Var=var)))

summarise(data.1, across(where(is.numeric), list(Mean=mean, Var=var)))

summarise(data.1, across(where(is.numeric), mean), across(where(is.factor), length)) ## with logical argument

```

Piping
=======

Grouping (=aggregating)
=========================

```{r}
data.1 %>% group_by(Between,Plot) %>% summarise(Mean=mean(Temp))

data.1 %>% group_by(Plot) %>% arrange(Temp) ##group each plot by temp

```
Reshaping data
=================
```{r}
## Wide data to long data - Gathering

data.w %>% gather(Time, Count, Time.0:Time.2)
```

```{r}
## Long data to wide - Spreading


```

Combining data
=================

```{r}
## ONLY KEEP FULL MATCHES BY BOTH - S3 and S7 will be absent

inner_join(data.bio, data.chem)

##other option is a full joint

full_join(data.bio, data.chem)

##left join keeps all of your first data set so S3 is missing and S7 is kept with NA for chem data
## Right join is reverse on it

left_join(data.bio, data.chem)
right_join(data.bio, data.chem)
## incorporate lat and longs into the bio data

left_join(data.bio, data.geo, by=c("Plot"))
```

Additional remarks on data wrangling

```{r}
library(tidyverse)

##playing with dates
temporal.data <- read.csv(file="AIMS R Course - Tutorial/data/temporal.data.csv", strip.white = TRUE)

## make dates into dates in R

temporal.data %>% mutate(Time=ymd_hms(Time))

date1 <- '2020-02-29'
date2 <- '2020-05-03'

ymd(date1)

as.numeric(ymd(date1))
18321
Diff=ymd(date2) - ymd(date1)
Diff
64

## prints it straight away if wrapped in brackets
(Diff=ymd(date2) - ymd(date1))

## express a time period
period(1, 'month')
ymd(date1) + period(1, 'month')
duration(1, 'month')

ymd(date1) + duration(1, 'month')

##aggregate dates - summary of times

temporal.sum <- temporal.data %>% mutate(Time = ymd_hms(Time)) %>% group_by(Site) %>% mutate(Hours=(Time - min(Time))/dhours(1), Hours = ceiling(Hours + 0.001)) %>% group_by(Site, Hours) %>% summarize(Mean=mean(Resp)) %>% ungroup()

##always ungroup at end (at least most of the time) because it does calculations based on grouping from here on

## Could do this for bull sharks, mean detections per month per bull shark per site


```

Applied examples
===================
