---
title: "GLMM example 5"
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
    css: ../resources/style.css
  pdf_document:
    df_print: default
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  word_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    toc: yes
    toc_depth: 2
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
---

```{r setup, include=FALSE, warnings=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparations
 
Load the necessary libraries

```{r libraries, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}
library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(ggfortify) #for model diagnostics
library(sjPlot)    #for outputs
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(MuMIn)     #for AICc
library(tidyverse) #for data wrangling
library(DHARMa)    #for assessing dispersion etc
```

# Scenario

Someone did something for some reason.....

# Read in the data

```{r readData, results='markdown', eval=TRUE}
owls = read_csv('../data/owls.csv', trim_ws=TRUE)
glimpse(owls)
```


# Exploratory data analysis
```{r}
owls <- owls %>% mutate(Nest=factor(Nest), FoodTreatment= factor(FoodTreatment), SexParent=factor(SexParent),Ncalls=SiblingNegotiation)

## start your models simple and built them up

ggplot(data = owls, aes(y=Ncalls, x = ArrivalTime, color=SexParent)) + geom_point() + geom_smooth(method='lm') + facet_grid(~FoodTreatment)

##lots of zeros in satiated group, not sure if we would expect as many 0s from a poisson. otherwise trends harder to tell, bit of an affect for arrival time and food treatment, parentsex more dispersed.

ggplot(data = owls, aes(y=Ncalls, x = BroodSize, color=SexParent)) + geom_point() + geom_smooth(method='lm') + facet_grid(~FoodTreatment)

## its very important that the predictors that you put in a model should have the same range of values. they need to overlap alot in x values

```


Model formula:
$$
y_i \sim{} \mathcal{Pois}(\lambda_i)\\
ln(\lambda_i) =\boldsymbol{\beta} \bf{X_i} + \boldsymbol{\gamma} \bf{Z_i}
$$

where $\boldsymbol{\beta}$ and $\boldsymbol{\gamma}$ are vectors of the fixed and random effects parameters respectively 
and $\bf{X}$ is the model matrix representing the overall intercept and effects of food treatment, sex of parent, arrival time (and various interactions) on the number of sibling negotiations.
Brood size was also incorporated as an offset.
$\bf{Z}$ represents a cell means model matrix for the random intercepts associated with individual nests.

# Fit the model
```{r}
## We have to account for the blocks in the data, random effects. Pick the things were actually interested in, mainly foodtreatment

owls.glmmTB <- glmmTMB(Ncalls ~ FoodTreatment + (1|Nest), data=owls, family=poisson(link='log'), REML=FALSE)

owls.glmmTB1 <- glmmTMB(Ncalls ~ FoodTreatment*scale(ArrivalTime) + (1|Nest), data=owls, family=poisson(link='log'), REML=FALSE)

## scale arrival time because it is continous

## add model that has additive form not interaction
owls.glmmTB2 <- glmmTMB(Ncalls ~ FoodTreatment + scale(ArrivalTime) + (1|Nest), data=owls, family=poisson(link='log'), REML=FALSE)

## this adds more complexity, 3 way interaction with parents
owls.glmmTB3 <- glmmTMB(Ncalls ~ SexParent*FoodTreatment*scale(ArrivalTime) + (1|Nest), data=owls, family=poisson(link='log'), REML=FALSE)

## lets check AICc
AICc(owls.glmmTB, owls.glmmTB1, owls.glmmTB2, owls.glmmTB3)

## besides the first only food, the other ones are similiar
#             df     AICc
#owls.glmmTB   3 5214.825
#owls.glmmTB1  5 5012.187
#owls.glmmTB2  4 5010.160
#owls.glmmTB3  9 5011.510


## lets try to standardize for the brood size, log it, offset is not modelled, just standardized (days at liberty in sharks), log it to remain as a 1:1 relationship, for a slope of one, needs to be on same scale

owls.glmmTB4 <- glmmTMB(Ncalls ~ FoodTreatment + scale(ArrivalTime) + offset(log(BroodSize)) + (1|Nest), data=owls, family=poisson(link='log'), REML=FALSE)

AICc(owls.glmmTB, owls.glmmTB1, owls.glmmTB2, owls.glmmTB3, owls.glmmTB4)
#             df     AICc
#owls.glmmTB   3 5214.825
#owls.glmmTB1  5 5012.187
#owls.glmmTB2  4 5010.160
#owls.glmmTB3  9 5011.510
#owls.glmmTB4  4 5009.073

## this one is better

## Update, with reml for unbiased estimates, were going into random effects now
owls.glmmTB4a = update(owls.glmmTB4, REML=TRUE)


## this keeps everything but take away the random effect that was there (1|nest) and replace with a new one, now random intercept random slope model associated with arrival time
owls.glmmTB4b = update(owls.glmmTB4a, .~.-(1|Nest) + (scale(ArrivalTime)|Nest))

AICc(owls.glmmTB4a, owls.glmmTB4b)
## b is even better

##this keeps everything but take away the random effect that was there (1|nest) and replace with a new one, now random intercept random slope model associated with arrival time and food treatment
owls.glmmTB4c = update(owls.glmmTB4a, .~.-(1|Nest) + (FoodTreatment+scale(ArrivalTime)|Nest))

AICc(owls.glmmTB4a, owls.glmmTB4b, owls.glmmTB4c)

## c is even better
## Further diagnostics

simulatedResiduals <- simulateResiduals(owls.glmmTB4c, plot=TRUE)

## it doesnt fit our nominated distribution, the lower end (tail) is bigger, more mass at lower end, than expected for a poisson. Is it because the model is overdispersed? Does not really appear to be. Could be corrected by negative binomial.There are outliers, some extra zeros. Graph on right looks at different parts of residuals (quantiles), suggest there is a bit of wobblinies but not the type that concerns us too much. if It was a parabula it would be concerning, our continous variable is arrival time, it might imply there is not a linear relationship with arrival time. the Bumps maybe are real, need more knowledge on ecology. We should ignore it, it is noise in residuals.

## is the model 0 inflated?
testZeroInflation(owls.glmmTB4c)
## this models a distribution of expected 0 from our data, our red line is way out of the distribution. We are certain it is 0 inflated.

testDispersion(owls.glmmTB4c)
## it is well within the usual simulated range, so not overdispersed

## now test autocorrelation

testTemporalAutocorrelation(owls.glmmTB4c)


## Now we will fit a 0 inflated model

owls.glmmTMB5 <- update(owls.glmmTB4c, ziformula=~1)
## tilt and then a 1, this is a model that only has an intercept in the 0 inflated part
## 0 inflation model fits the two processes underlying our data (0 could be artefact of simply not detecting calls, model just accounts for the fact that I did not detect any calls)

summary(owls.glmmTMB5)
## pay attention first to the 0 inflated part
## 0 inflated model is binomial on log odds scale
exp(-1.3705)
##0.25 is probability of detecting, of our 0s 4 times more of them are considered falls 0 than actual true 0s
## Maybe we should investigate our 0s. Are they related to something? food treatment or arrival time?

owls.glmmTMB6 <- update(owls.glmmTB4c, ziformula=~scale(ArrivalTime))
summary(owls.glmmTMB6)
## parallel lines = no exception
exp(0.75177)
#2.12075 = y when x = 0 for the deprived group (intercept) average number of calls per chick for the deprived group at the average arrival time.
exp(-0.75562)
#0.4697193, same position for satiated group
#exp(0.75177-0.75562) = 0.99, on average satiated group only has one call per chick at average arrival time, it declined by 1-0.469 = is a 53% decline.
exp(-0.16517)
##0.8477496 for every 1 unit change in scaled arrival time, the number of calls declines by 16%

## R^2 value

r.squaredGLMM(owls.glmmTMB6)
#ERROR cannot yet handle object with zero inflation

##with means that you dont have keep writing owl$ this owl$ that is sets to scope variables are always in owl
owls.grid <- with(owls, list(FoodTreatment=levels(FoodTreatment), ArrivalTime=modelr::seq_range(ArrivalTime, n=100), SexParent=levels(SexParent)))

## reminder, its not gonna use sex parent as it was not in the model

newdata = emmeans(owls.glmmTMB6, ~ArrivalTime|FoodTreatment, at=owls.grid, type='response') %>% as.data.frame()

ggplot(newdata, aes(y=rate, x=ArrivalTime)) +
  geom_line(aes(color=FoodTreatment)) +
  geom_ribbon(aes(ymin=lower.CL, ymax=upper.CL, fill=FoodTreatment), alpha=0.2)


## if u put raw data back on graph it wouldnt match it wouldnt account for 0 inflation, it wouldnt account for brood size, it wouldnt account for blocks, 
```


# Model validation

# Model investigation / hypothesis testing

# Predictions

# Summary figures

# References
