---
title: "GLMM example 4"
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
    css: ../resources/style.css
  pdf_document:
    df_print: default
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  word_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    toc: yes
    toc_depth: 2
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
---

```{r setup, include=FALSE, warnings=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparations

Load the necessary libraries

```{r libraries, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}
library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(ggfortify) #for model diagnostics
library(sjPlot)    #for outputs
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(MuMIn)     #for AICc
library(tidyverse) #for data wrangling
```

# Scenario

Someone did something for some reason.....

# Read in the data

```{r readData, results='markdown', eval=TRUE}
mckeon = read_csv('../data/mckeon.csv', trim_ws=TRUE)
glimpse(mckeon)
```




Model formula:
$$
y_i \sim{} \mathcal{N}(n, p_i)\\
ln\left(\frac{p_i}{1-p_1}\right) =\boldsymbol{\beta} \bf{X_i} + \boldsymbol{\gamma} \bf{Z_i}
$$

where $\boldsymbol{\beta}$ and $\boldsymbol{\gamma}$ are vectors of the fixed and random effects parameters respectively 
and $\bf{X}$ is the model matrix representing the overall intercept and effects of symbionts on the probability of the colony experiencing predation.
$\bf{Z}$ represents a cell means model matrix for the random intercepts associated with individual coral colonies.

# Exploratory data analysis and # Fit the model # Model validation
```{r}
## Predation on corals (single factor anova), treatment is wether having a symbiont in the form of a crab or shrimp can reduce predation. Blocks are tanks, the symbiont is the treatment. Each treatment is applied to each block a randomized block design. 
## We should also change the order of symbionts because its alphabetical now for the model, as we declare our factors lets change order of levels for symbionts

mckeon <- mckeon %>% mutate(BLOCK=factor(BLOCK), SYMBIONT=factor(SYMBIONT, levels=c('none','crabs','shrimp','both')))

##Plot: X axis we want factor with most levels (symbiont), y-axis is response (predation 0 or 1) and plot it seperately for each plot so we can see how each one of them performs 

ggplot(mckeon, aes(y=PREDATION, x=SYMBIONT)) + geom_point(position=position_jitter(width=0.2, height=0)) + facet_wrap(~BLOCK)

## Does not look like baseline likelihood for predation (intercept) is not the same. Random intercepts vs. random slopes
## We know were using a binomial distribution because we only have 0s and 1s
#Lets fit a random intercept and then a random slope model

mckeon.glmmTMB <- glmmTMB(PREDATION ~ SYMBIONT + (1|BLOCK), data=mckeon, family=binomial(link='logit'), REML=FALSE)

mckeon.glmmTMB1 <- glmmTMB(PREDATION ~ SYMBIONT + (SYMBIONT|BLOCK), data=mckeon, family=binomial(link='logit'), REML=FALSE)

## did not work, rerun with REML

mckeon.glmmTMB <- update(mckeon.glmmTMB, REML=TRUE)

plot_model(mckeon.glmmTMB, type='diag')
##ggplot plot looks good, but binomials are harder to interpret so we only get one graph but have learned a new method that gives us more info, graph on right shows the symbionts (1-4), there residuals

simulateResiduals <- simulateResiduals(mckeon.glmmTMB, plot=TRUE)
## these diagnostics look good

summary(mckeon.glmmTMB)

## units for means are on a log odds scale, 4.42 for the non group
exp(4.42)
## 83.09, being predated is 83% higher than not being predated on in the none group
##crab group: the odds of predation are substantially less than the none group, risk of predation is only 4% of that, a 96% decline in odds of being predated. They are still experiencing predation but not as bad as none group
## what about both group
exp(-5.15)
## is 0.005799, odds of 0.5 to be predate, less likely to be predated than predated, 99.4% reduction on odds being predated.

tidy(mckeon.glmmTMB, effect='fixed', conf.int=TRUE, exponentiate = TRUE) %>% kable
## This only tells us every info in relation to none
## If we want to know crab vs. shrimp, or either 1 vs. having 2, or comparing none, vs. having a symbiont, we need a contrast matrix, we reparameteriz our model to fit out specified categories.
#         crab vs. shrimp      1 vs 2      none vs symbiont
# none        0                   0                1
# crabs       1                  -0.5              -1/3
# shrimp      -1                 -0.5              -1/3
# both         0                   1               -1/3
# can make 3 comparisons for 4 groups, but need be independent

cmat=cbind(crab_vs_shrimp=c(0,1,-1,0), one_vs_both=c(0,-1/2,-1/2,1), none_vs_symbiont=c(1,-1/3,-1/3,-1/3))
round(crossprod(cmat),1)

emmeans(mckeon.glmmTMB, ~SYMBIONT, contr=list(SYMBIONT=cmat), type='response')

## the odds of predation are nearly 90% higher for crabs than if there are shrimp (but non-significant)
## one vs. both, non-significant, symbiont vs. none is significant
##e.g. comparing 1vs2 symbionts 1/0.22 = 4.54, you are 4.5 times more likely to be predated on with 1 instead of two

## R^2
r.squaredGLMM(mckeon.glmmTMB)
#                  R2m       R2c
#theoretical 0.1489321 0.8674549
#delta       0.1437647 0.8373578
# when we account for the treatments we can explain about 14% of the variability, it does have an effect but its only explaining 14%. When we put the random blocks(tanks) in, we can explain 84% of the variability. There is quiet some difference between the tanks themselves. 



## Summary plot

emmeans(mckeon.glmmTMB, ~SYMBIONT, type='response') %>% as.data.frame() %>%
  ggplot(aes(y=prob, x=SYMBIONT)) +
  geom_pointrange(aes(ymin=lower.CL, ymax=upper.CL))
```


```{r}

```


